Shallow EDSLs and Object-Oriented Programming

### 2016-06-30 paper/paper.tex
url + client code
4 pages
More rewritting for intro

\setlist[itemize]{noitemsep,nolistsep}
\setlist[enumerate]{noitemsep,nolistsep}

@@

Shallow Embedded Domain Specific Languages (EDSLs) use
\emph{procedural abstraction} to directly encode a DSL into an existing host language. Procedural abstraction has
been argued to be the essence of Object-Oriented Programming (OOP). 
%Given
%that OO languages have evolved over more than 50 years
%to improve the use of procedural abstraction, they ought to have some
%advantages to encode shallow EDSLs.
This paper argues that OOP abstractions 
(including \emph{inheritance}, \emph{subtyping} and
\emph{type-refinement}) 
increase the modularity and reuse of shallow
EDSLs when compared to classical procedural abstraction. We make this
argument by taking a recent paper by Gibbons and Wu, where procedural
abstraction is used in Haskell to model a simple EDSL, and we recode
that EDSL in Scala. From the \emph{semantic}
and \emph{modularity} point of view the Scala version has clear advantages 
over the Haskell version.

>>>>

\setlist[itemize]{noitemsep,nolistsep}
\setlist[enumerate]{noitemsep,nolistsep}
\setlength{\floatsep}{6.0pt plus 2.0pt minus 2.0pt}
\setlength{\textfloatsep}{10.0pt plus 2.0pt minus 4.0pt}
\setlength{\intextsep}{6.0pt plus 2.0pt minus 2.0pt}

@@

Shallow Embedded Domain Specific Languages (EDSLs) use
\emph{procedural abstraction} to directly encode a DSL into an existing host language. Procedural abstraction has
been argued to be the essence of Object-Oriented Programming (OOP). 
%Given
%that OO languages have evolved over more than 50 years
%to improve the use of procedural abstraction, they ought to have some
%advantages to encode shallow EDSLs.
This paper argues that OOP abstractions 
(including \emph{inheritance}, \emph{subtyping} and
\emph{type-refinement}) 
increase the modularity and reuse of shallow
EDSLs when compared to classical procedural abstraction. We make this
argument by taking a recent paper by Gibbons and Wu, where procedural
abstraction is used in Haskell to model a simple shallow EDSL, and we recode
that EDSL in Scala. From the \emph{semantic}
and \emph{modularity} point of view the Scala version has clear advantages 
over the Haskell version.

### 2016-06-30 paper/sections/Introduction.tex
More rewritting for intro

\noindent Although the above definition is quite reasonable and widely accepted,
it leaves some space to (mis)interpretation. For example it is unclear 
how to classify an EDSL implemented using the {\sc Composite} or {\sc Interpreter} 
patterns in Object-Oriented Programming (OOP). Would this OO approach be
classified as a shallow or deep embedding? Some authors working on
OOP EDSLs ~\cite{} consider a {\sc Composite} to be a deep
embedding. Some other authors~\cite{gibbons2014folding} 
consider implementations using tuples and/or the  {\sc Composite}
pattern to be a shallow embedding.

In this paper, we avoid ambiguitity by using a more precise meaning
for shallow embeddings as EDSLs implemented using \emph{procedural abstraction}~\cite{reynolds94proceduralabstraction}. Such
interpretation arises naturally from the domain of shallow EDSLs being
typically functions, and procedural abstraction being a way to encode
data abstractions using functions. As Cook~\cite{cook09abstraction} argued,
procedural abstraction is also the essence of OOP.
Thus, according to our definition, the implementation of a shallow
EDSL in OOP languages should simply correspond to a standard
object-oriented program.

The main goal of our paper is to show that OOP languages have
advantages for the implementation of shallow embeddings. 
It is perhaps due to those advantages that some authors 
have considered  {\sc Composite}-based implementations to be deep
embeddings, rather than shallow. However, we argue that the OOP
mechanisms do not change the essence of the implementation, which is
still shallow (i.e. using procedural abstraction). To understand how
OOP is helpful for shallow embeddings, we should first look at the
limitations of shallow embeddings commonly found in the literature:
\begin{enumerate}

\item {\bf Single Interpretation} An often stated limitation of
  shallow embeddings is that they only support a single
  interpretation.

\item {\bf No Transformations} Another commonly stated limitation 
of shallow embeddings is that they do not support transformations,
preventing optimizations and other useful transformations.

@@

We make our arguments by taking a recent paper by Gibbons and Wu~\cite{gibbons2014folding},
where procedural abstraction is used in Haskell to model a simple
EDSL, and we recode that EDSL in Scala\footnote{Available online: \url{https://github.com/wxzh/shallow-dsl}}.
From the \emph{modularity} point of view the
Scala version has clear advantages over the Haskell version.

>>>>

\noindent Although the above definition is quite reasonable and widely accepted,
it leaves some space to (mis)interpretation. For example it is unclear 
how to classify an EDSL implemented using the {\sc Composite} or {\sc Interpreter} 
patterns in Object-Oriented Programming (OOP). Would this OO approach be
classified as a shallow or deep embedding? We feel there is a rather
fuzzy line here, and the literature allows for both interpretations. Some authors working on
OOP EDSLs ~\cite{} consider a {\sc Composite} to be a deep
embedding. Some other authors~\cite{gibbons2014folding} 
consider implementations using tuples and/or the  {\sc Composite}
pattern to be a shallow embedding.

To avoid ambiguitity we propose defining shallow embeddings as EDSLs implemented using \emph{procedural abstraction}~\cite{reynolds94proceduralabstraction}. Such
interpretation arises naturally from the domain of shallow EDSLs being
functions, and procedural abstraction being a way to encode
data abstractions using functions. As Cook~\cite{cook09abstraction} argued,
procedural abstraction is also the essence of OOP.
Thus, according to our definition, the implementation of a shallow
EDSL in OOP languages should simply correspond to a standard
object-oriented program.

The main goal of our paper is to show the close relationship between 
shallow embeddings and OOP, and argue that OOP languages have
advantages for the implementation of shallow embeddings. 
It is perhaps partly due to those advantages that some authors 
have considered  {\sc Composite}-based implementations to be deep
embeddings. However, we argue that the OOP
mechanisms do not change the essence of the implementation, which is
still shallow (i.e. using procedural abstraction). 
To understand how
OOP is helpful for shallow embeddings, we should first look at the
limitations of shallow embeddings commonly found in the literature:
\begin{enumerate}

\item {\bf Single Interpretation} An often stated limitation of
  shallow EDSLs is that they only support one
  interpretation.

\item {\bf No Transformations} Another commonly stated limitation 
of shallow embeddings is the lack of support for transformations,
preventing optimizations.

@@

We make our arguments by taking a recent paper by Gibbons and Wu~\cite{gibbons2014folding},
where procedural abstraction is used in Haskell to model a simple \emph{shallow}
EDSL, and we recode that EDSL in Scala\footnote{Available online: \url{https://github.com/wxzh/shallow-dsl}}.
From the \emph{modularity} point of view the
Scala version has clear advantages over the Haskell version.
